---
title: "Dataframes - In Class lab"
author: "Claudia Dominguez"
date: "DATA 2401"
output: "html_document"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Working with loaded in csv files


Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants` 

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")

head(grants) 
str(grants)   
summary(grants)  


```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

- Use the glimpse function to look at the loaded data

```{r}
getwd()

library(dplyr)  

glimpse(grants) 


```


- Create a variable `org` that contains the `organization` column of the dataset

```{r}
library(dplyr)

org <- grants %>% pull(organization)

head(org)


```


- Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}

is.vector(org)


```


Now you can ask some interesting questions about the dataset:

- What was the mean grant value?

```{r}
mean_grant <- mean(grants$total_amount, na.rm = TRUE)
mean_grant

```


- What was the dollar amount of the largest grant?

```{r}
largest_grant <- max(grants$total_amount, na.rm = TRUE)
largest_grant

```


- What was the dollar amount of the smallest grant?

```{r}

smallest_grant <- min(grants$total_amount, na.rm = TRUE)
smallest_grant

```


- Which organization received the largest grant?

```{r}

largest_grant_index <- which.max(grants$total_amount)


largest_grant_org <- grants$organization[largest_grant_index]


largest_grant_org


```


- Which organization received the smallest grant?

```{r}

smallest_grant_index <- which.min(grants$total_amount)


smallest_grant_org <- grants$organization[smallest_grant_index]


smallest_grant_org

```

- How many grants were awarded in 2010?

```{r}
colnames(grants)

grants_2010_count <- sum(grants$start_year == 2010, na.rm = TRUE)

grants_2010_count




```

- How much was awarded in 2010? 

```{r}

total_awarded_2010 <- sum(grants$total_amount[grants$start_year == 2010], na.rm = TRUE)

total_awarded_2010

```


# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`
```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
expenditure_df <- data.frame(USPersonalExpenditure)

str(expenditure_df)

head(expenditure_df)

```


What are the current column names of your dataframe?
```{r}

colnames(expenditure_df)

```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}

rownames(expenditure_df)

```


Add a column "Category" to your data frame that contains the rownames

```{r}
expenditure_df$Category <- rownames(expenditure_df)

head(expenditure_df)

```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}
library(tidyr)

tidy_expenditure <- expenditure_df %>%
  pivot_longer(
    cols = -Category,  
    names_to = "Year",  
    values_to = "Expenditure"  
  )

head(tidy_expenditure)

```


How much money was spent on personal care in 1940?
```{r}

dim(expenditure_df)


head(expenditure_df)


colnames(expenditure_df)



# I gave up i don't know how to do the next 5 questions




```


How much money was spent on Food and Tobacco in 1960?
```{r}
library(dplyr)

food_tobacco_1960 <- tidy_expenditure %>%
  filter(Category == "Food and Tobacco" & Year == "1960") %>%
  select(Expenditure)

food_tobacco_1960


```


What was the highest expenditure category in 1960?
```{r}
library(dplyr)

highest_expenditure_1960 <- tidy_expenditure %>%
  filter(Year == "1960") %>%
  arrange(desc(Expenditure)) %>%
  slice(1)  

highest_expenditure_1960


```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}

library(dplyr)

lowest_category <- function(year) {
  result <- tidy_expenditure %>%
    filter(Year == as.character(year)) %>%  
    arrange(Expenditure) %>%  
    slice(1)  

  return(result)
}





```

```{r}

lowest_category(1960)
```


Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}
years <- unique(tidy_expenditure$Year)

lowest_categories_per_year <- sapply(years, function(y) lowest_category(y)$Category)

lowest_categories_per_year

```
